### 2023-07-06 paper/localpsf.tex
removing less important references to shorten paper

We present an efficient \emph{matrix-free} point spread function (PSF) method for approximating operators $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, and $L^2(\Omega)'$ is the space of real-valued continuous linear functionals on $L^2(\Omega)$. Such operators appear, for instance, as Hessians in optimization and inverse problems governed by partial differential equations (PDEs)~\cite{AntilEtAl18,BorziSchulz11,Delosreyes15,HinzePinnauUlbrichUlbrich08}, Schur complements in Schur complement methods for solving partial differential equations (PDEs) and Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps)~\cite{ChanMathew94,SmithBjorstadGropp96,ToselliWidlund04}, covariance operators in spatial statistics~\cite{ChenStein21,GentonKeyesTurkiyyah18,GeogaEtAl20,LindgrenRueLindstrom11}, and blurring operators in imaging~\cite{DenisEtAl11,NagyOleary98}. Here ``matrix-free'' means that we may apply $\Aop$ and its transpose\footnote{Recall that $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop^T w\right)(u)$ for all $u,w \in L^2(\Omega)$, where $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for operations with $\Aop^T$. }, $\Aop^T$, to functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
via a black box computational procedure, 
but cannot easily access entries of $\Aop$'s integral kernel. Evaluating the maps in~\eqref{eq:eval_maps} may require solving a subproblem that involves PDEs, or performing other costly computations. By ``non-negative integral kernel,'' we mean that entries of the integral kernel are non-negative numbers; this is not the same as positive semi-definiteness of $\Aop$.

@@

\paragraph{Lumped mass matrix} At the continuum level, $\Aker$ is assumed to be non-negative. However, entries of $\boldsymbol{\Aker}$ involve inverse mass matrices, which typically contain negative numbers. 
We therefore recommend replacing the mass matrix, $\massmatrix$, with a positive diagonal \emph{lumped mass} approximation. 
Here, we use the lumped mass matrix in which the $i$th diagonal entry of the lumped mass matrix is the sum of all entries in the $i$th row of the mass matrix. Other mass lumping techniques may be used~\cite{Hughes12}.

@@

Let $x$ and $x'$ be points in $\Omega$ that are close to each other, and consider the following approximations:
\begin{align}
\phi_x(y) &\approx \phi_{x'}(y) \label{eq:translate1}\\
\phi_x(y) &\approx \phi_{x'}(y-x+x') \label{eq:translate2}\\
\phi_x(y) &\approx \phi_{x'}\left(y-\spatialmean(x)+\spatialmean(x')\right) \label{eq:translate3}\\
\phi_x(y) &\approx \phi_{x'}\left(y-\spatialmean(x)+\spatialmean(x')\right) \spatialvol(x) / \spatialvol(x'). \label{eq:translate4}
\end{align}
These are four different ways to approximate an impulse response by a nearby impulse response, with each successive approximation building upon the previous ones. The proposed method uses~\eqref{eq:translate4}, which is the most sophisticated. Approximation~\eqref{eq:translate1} says that $\phi_x$ can be approximated by $\phi_{x'}$ when $x$ and $x'$ are close. Operators satisfying~\eqref{eq:translate1} can be well approximated via low rank CUR approximation~\cite{MahoneyDrineas09}. However, the required rank in the low rank approximation can be large, which makes algorithms based on~\eqref{eq:translate1} expensive. Operators that satisfy~\eqref{eq:translate2} are called ``locally translation invariant'' because integral kernel entries $\Aker(y,x)$ for such operators are approximately invariant under translation of $x$ and $y$ by the same displacement, i.e., $x \rightarrow x+h$ and $y \rightarrow y+h$. It is straightforward to show that if equality holds in~\eqref{eq:translate2}, then $\Aop$ is a convolution operator. Locally translation invariant operators act like convolutions locally, and can therefore be well approximated by PC approximations.

>>>>

We present an efficient \emph{matrix-free} point spread function (PSF) method for approximating operators $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, and $L^2(\Omega)'$ is the space of real-valued continuous linear functionals on $L^2(\Omega)$. Such operators appear, for instance, as Hessians in optimization and inverse problems governed by partial differential equations (PDEs)~\cite{BorziSchulz11,Delosreyes15,HinzePinnauUlbrichUlbrich08},
%AntilEtAl18
Schur complements in Schur complement methods for solving partial differential equations (PDEs) and Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps)~\cite{ChanMathew94,SmithBjorstadGropp96,ToselliWidlund04}, covariance operators in spatial statistics~\cite{ChenStein21,GentonKeyesTurkiyyah18,GeogaEtAl20,LindgrenRueLindstrom11}, and blurring operators in imaging~\cite{DenisEtAl11,NagyOleary98}. Here ``matrix-free'' means that we may apply $\Aop$ and its transpose\footnote{Recall that $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop^T w\right)(u)$ for all $u,w \in L^2(\Omega)$, where $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for operations with $\Aop^T$. }, $\Aop^T$, to functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
via a black box computational procedure, 
but cannot easily access entries of $\Aop$'s integral kernel. Evaluating the maps in~\eqref{eq:eval_maps} may require solving a subproblem that involves PDEs, or performing other costly computations. By ``non-negative integral kernel,'' we mean that entries of the integral kernel are non-negative numbers; this is not the same as positive semi-definiteness of $\Aop$.

@@

\paragraph{Lumped mass matrix} At the continuum level, $\Aker$ is assumed to be non-negative. However, entries of $\boldsymbol{\Aker}$ involve inverse mass matrices, which typically contain negative numbers. 
We therefore recommend replacing the mass matrix, $\massmatrix$, with a positive diagonal \emph{lumped mass} approximation. 
Here, we use the lumped mass matrix in which the $i$th diagonal entry of the lumped mass matrix is the sum of all entries in the $i$th row of the mass matrix. Other mass lumping techniques may be used.
%~\cite{Hughes12}.

@@

Let $x$ and $x'$ be points in $\Omega$ that are close to each other, and consider the following approximations:
\begin{align}
\phi_x(y) &\approx \phi_{x'}(y) \label{eq:translate1}\\
\phi_x(y) &\approx \phi_{x'}(y-x+x') \label{eq:translate2}\\
\phi_x(y) &\approx \phi_{x'}\left(y-\spatialmean(x)+\spatialmean(x')\right) \label{eq:translate3}\\
\phi_x(y) &\approx \phi_{x'}\left(y-\spatialmean(x)+\spatialmean(x')\right) \spatialvol(x) / \spatialvol(x'). \label{eq:translate4}
\end{align}
These are four different ways to approximate an impulse response by a nearby impulse response, with each successive approximation building upon the previous ones. The proposed method uses~\eqref{eq:translate4}, which is the most sophisticated. Approximation~\eqref{eq:translate1} says that $\phi_x$ can be approximated by $\phi_{x'}$ when $x$ and $x'$ are close. Operators satisfying~\eqref{eq:translate1} can be well approximated via low rank CUR approximation.
%~\cite{MahoneyDrineas09}.
However, the required rank in the low rank approximation can be large, which makes algorithms based on~\eqref{eq:translate1} expensive. Operators that satisfy~\eqref{eq:translate2} are called ``locally translation invariant'' because integral kernel entries $\Aker(y,x)$ for such operators are approximately invariant under translation of $x$ and $y$ by the same displacement, i.e., $x \rightarrow x+h$ and $y \rightarrow y+h$. It is straightforward to show that if equality holds in~\eqref{eq:translate2}, then $\Aop$ is a convolution operator. Locally translation invariant operators act like convolutions locally, and can therefore be well approximated by PC approximations.
